Co-methylation

### 2014-12-22 Rmarkdown/comethylation.Rmd
Discussion of LORs

__TODO: How often are points in these plots from different CGIs? Compute distribution of CGI widths and gaps between them__

@@

__TODO: Periodicities__
- The periodicities (10? and 167ish) are more noticeable in the non-island pairs than the pairs in CpG islands. Possible reasons?
- A shorter periodicity over the first 100 bp?

@@

Recall that for each 2-tuple and sample we can summarise the methylation patterns by a $2 \times 2$ contingency table (__FIGURE__). The within-fragment co-methylation of the pair of CpGs, $(i, i')$, in the $j^{th}$ sample is then defined as the logarithm of the odds ratio of this table, $LOR_{(i, i'), j} = log_2 \Big (\frac{MM_{i, i'} \times UU_{i, i'}}{MU_{i, i'} \times UM_{i, i'}} \Big)$. I use base-2 logarithms due to their widespread use and familiarity within biology. In the `cometh` function I also add a small offset (default `offset = 1`) to avoid problems arising from zero counts, which frequently occur.

A positive $LOR$ is evidence that the methylation states are likely to be the same at both CpGs in the pair on the same DNA fragment, a negative $LOR$ is evidence that the methylation states are likely to be different at each CpG in the pair on the same DNA fragment, and a value close to zero is evidence that the methylation state at each CpG in the pair on the same DNA fragment are independent of one another. More precisely, we may wish to compute a confidence interval for $LOR$ or perform a hypothesis test to determine whether the observed $LOR$ is evidence for co-methylation at this pair of CpGs.

@@

@@

- LOR decreases with IPD
- For a given IPD, a wide variation in LOR
- LOR does not go to zero over observable distance
- Variation is fairly constant across IPD
- Noise at the end is due to limited data
- CGIs are more co-methylated than non-CGI
- SE data can't cut the mustard
- Intriguing mode/plateua around 150-200 bp (NIL >= 0)
  - SE data suggests some is technical, e.g., incomplete removal of M-bias
  - Also, IMR90 PE data doesn't have it
- LOR vs. Pearson
- NIL = 0 vs. NIL >= 0
- iPSC vs. somatic
- Plot histogram of LOR and histogram of r
- Look for tables with LOR > 0 and r = 0

@@

>>>>

@@

The $150-200$ bp periodicity is also evident from the $NIL \geq 0$ data and is particularly clear for the CpG pairs outside of CpG islands. __TODO: Discuss more__

@@

Recall that for each 2-tuple and sample we can summarise the methylation patterns by a $2 \times 2$ contingency table, like that shown in Table \ref{tab:2x2}. The within-fragment co-methylation of the pair of CpGs, $(i, i')$, in the $j^{th}$ sample is then defined as the logarithm of the odds ratio of this table, $LOR_{(i, i'), j} = \log_{2} \Big (\frac{mm_{i, i'} \times uu_{i, i'}}{mu_{i, i'} \times um_{i, i'}} \Big)$. I use base-2 logarithms due to their widespread use and familiarity within biology. In the `cometh` function I also add a small offset (default `offset = 1`) to avoid problems arising from zero counts, which frequently occur. For the example in Table \ref{tab:2x2} with `offset = 1` we have $LOR = \log_{2} \Big (\frac{(1 + 1) \times (4 + 1)}{(0 + 1) \times (4 + 1)} \Big) = 1.74$.

From a Bayesian perspective, the offset may be interpreted as a prior count from a table with no association, i.e., a table with $LOR = 0$. The offset then has the effect of shrinking the raw estimate of the $LOR$, with more shrinkage applied for larger values of `offset`. Analyses using different offsets should be compared with caution.

\begin{table}[h]
\centering
\begin{tabular}{lcccl}
                           & \multicolumn{1}{l}{} & \multicolumn{2}{l}{Second CpG} &                \\
                           &                      & \textbf{M}     & \textbf{U}    & \textbf{Total} \\ \cline{2-5}
\multirow{2}{*}{First CpG} & \textbf{M}           & 1              & 2             & 3              \\
                           & \textbf{U}           & 0              & 4             & 4              \\ \cline{2-5}
                           & \textbf{Total}       & 1              & 6             & 7
\end{tabular}
\caption{Example $2 \times 2$ contingency table summarising methylation patterns at a 2-tuple.}
\label{tab:2x2}
\end{table}

A positive $LOR$ is evidence that the methylation states of two CpGs on the same DNA fragment are likely to be the identical, a negative $LOR$ is evidence that the methylation states of two CpGs on the same DNA fragment are likely to be the different, and a value close to zero is evidence that the methylation states of two CpGs on the same DNA fragment are independent of one another. More precisely, we may wish to compute a confidence interval for the $LOR$ or perform a hypothesis test to determine whether the observed $LOR$ is evidence for co-methylation at this pair of CpGs.

There are several alternative measures of association for a $2 \times 2$ table of binary variables such as __EXAMPLES__

@@

- Plot histogram of LOR and histogram of r

- Look for tables with LOR > 0 and r = 0 (and vice versa?)

@@

- LOR decreases with IPD
- For a given IPD, a wide variation in LOR
- LOR does not go to zero over observable distance
- Variation is fairly constant across IPD
- Noise at the end is due to limited data
- CGIs are more co-methylated than non-CGI
- SE data can't cut the mustard
- Intriguing mode/plateua around 150-200 bp (NIL >= 0)
  - SE data suggests some is technical, e.g., incomplete removal of M-bias
  - Also, IMR90 PE data doesn't have it
- LOR vs. Pearson
- NIL = 0 vs. NIL >= 0
- iPSC vs. somatic

@@

\section{Summary}

- Discuss 1D nature of $IPD$s vs. $4D$ structure of genome (3D + time)
